---
title: "Practical 2: Non-linear Regression"
author: "Garyfallos Konstantinoudis"
date: "Spring Term 2024"
output: pdf_document
highlight: tango
---

```{r include=FALSE}
knitr::opts_chunk$set(message=FALSE, warning=FALSE, fig.align = 'center')
```

```{r eval=FALSE}
devtools::install_github("ZheyuanLi/SplinesUtils")
install.packages(c("dplyr", "ggplot2", "patchwork", "splines", "blmeco", "mgcv", "dlnm",
                   "plotly"))
```
```{r}
library(dplyr)
library(ggplot2)
library(patchwork)
library(splines)
library(blmeco)
library(mgcv)
library(dlnm)
library(SplinesUtils)
library(plotly)
```


## Part 1: Triceps and subscapular skinfold thicknesses using basis functions

Triceps and subscapular skinfold thicknesses provide an index of body fat and midarm muscle circumference provides a measure of muscle mass. The dataset \texttt{triceps.csv} contains the age in years, the intriceps and triceps skinfold thickness in cm. In the first part of this practical we would like to examine the association between age and triceps skinfold thickness.

As a first step we load the data.

```{r eval=TRUE}
triceps <- read.csv("../../data/triceps.csv")
```

Question 1.1
Perform the \texttt{head()} function to the dataset to understand the data structure. Fit a regression model (Normal, Poisson or logistic depending on the nature of the main outcome) to examine the linear association between triceps skinfold thickness and age. Provide an interpretation of the result.

*Reply: For each additional year in the age there is a 0.2 (0.20, 0.24) cm increase in the triceps skinfold thickness*

```{r fig.width=4.8, fig.height=3.5}
triceps %>% head()
triceps %>% ggplot() + geom_histogram(aes(x=triceps)) + theme_bw()
lm(triceps~age, data = triceps) -> mod_linear
mod_linear %>% summary()
mod_linear %>% confint()
```


Question 1.2
What is the main assumption regarding the shape of the relationship between triceps skinfold thickness and age in question 1.1? Can you check if this assumption is valid?

*Reply: The main assumption is that the relationship between triceps skinfold thickness and age is linear. We can plot them in a scatterplot to see if this assumption is valid.*



*Having plotted the results, the main drawback of the approach in question 1.1 is that it misses the initial declining trend for the first 10 years of life.*

Question 1.3
Fit 3 models with polynomial basis function of order 2, 3 and 4. Plot the results together and discuss. Perform a series of likelihood ratio test to examine which model fits the data best.

```{r fig.width=6, fig.height=4}
mod_po2 <- lm(triceps~age+I(age^2), data = triceps)
mod_po3 <- lm(triceps~age+I(age^2)+I(age^3), data = triceps)
mod_po4 <- lm(triceps~age+I(age^2)+I(age^3)+I(age^4), data = triceps)

data.frame(x=rep(triceps$age, times = 3),
           fitted=c(mod_po2$fitted.values, mod_po3$fitted.values,
                    mod_po4$fitted.values),
           pol = rep(c("Order 2", "Order 3", "Order 4"),
                     each = mod_po2$fitted.values %>% length())) %>%
  ggplot() + geom_point(data = triceps, aes(x = age, y = triceps), cex = 1.2) +
  geom_line(aes(x = x, y = fitted, col = pol), cex = 1.2) +
  theme_bw()
```

*It is interesting that the mod_po2 gives zero coefficient to the $x^2$, with the effect being very similar with the linear approach.*
```{r}
mod_po2 %>% summary()
```

*and the likelihood ratio tests:*
```{r}
anova(mod_po2, mod_po3)
anova(mod_po3, mod_po4)
anova(mod_po2, mod_po4)
```

*The likelihood ratio test shows the need of a more flexible model, with degree 4 polynomial. *

Question 1.4
Select the best performing model of the question 1.3 and plot the flexible fit together with the 95% confidence intervals. Hint: Use the function \texttt{predict()} to get the standard error and compute the upper and lower limit of the confidence intervals.

*Reply:*
```{r fig.width=6, fig.height=4}
predict(mod_po4, newdata = data.frame(age = triceps$age), se.fit = TRUE) -> pred_p

data.frame(x = triceps$age,
           mean = pred_p$fit,
           UL = pred_p$fit + 1.96*pred_p$se.fit,
           LL = pred_p$fit - 1.96*pred_p$se.fit
           ) %>%
  ggplot() + geom_point(data = triceps, aes(x = age, y = triceps), cex = 1.2) +
  geom_line(aes(x = x, y = mean), cex = 1.2) +
  geom_ribbon(aes(x = x, ymin = LL, ymax = UL), fill = "blue", alpha = 0.2) +
  theme_bw()
```

Question 1.5
Now use a Fourier basis function and fit three models with one sin/cos pair and period $P=25,50,100$ years. Plot the results together and discuss.

*Reply:*
```{r fig.width=6.5, fig.height=4}
omega <- 2*pi/25
mod_fourier_p25 <- lm(triceps~I(sin(omega*age))+I(cos(omega*age)), data = triceps)
omega <- 2*pi/50
mod_fourier_p50 <- lm(triceps~I(sin(omega*age))+I(cos(omega*age)), data = triceps)
omega <- 2*pi/100
mod_fourier_p100 <- lm(triceps~I(sin(omega*age))+I(cos(omega*age)), data = triceps)


data.frame(x=rep(triceps$age, times = 3),
           fitted=c(mod_fourier_p25$fitted.values, mod_fourier_p50$fitted.values, mod_fourier_p100$fitted.values),
           pol = rep(c("P=25", "P=50", "P100"), each = mod_fourier_p25$fitted.values %>% length())) %>%
  ggplot() + geom_point(data = triceps, aes(x = age, y = triceps), cex = 1.2) +
  geom_line(aes(x = x, y = fitted, col = pol), cex = 1.2) +
  theme_bw()
```




## Part 2: Triceps and subscapular skinfold thicknesses with linear splines

Question 2.1
Use the same dataset as before and now fit 2 linear threshold models (not linear splines!), the first with one threshold at 10 years, and the second with two thresholds one at 10 and one at 40 years. Plot the results together and discuss. How do you interpret the results for the different thresholds? Discuss potential limitations of this approach.

*Reply:*
```{r}
mod_lt_1 <- lm(triceps~age:(age<10), data = triceps)
mod_lt_1 %>% summary()

predict(mod_lt_1, newdata = data.frame(age = triceps$age), se.fit = TRUE) -> pred_lt_1

data.frame(x = triceps$age,
           mean = pred_lt_1$fit,
           UL = pred_lt_1$fit + 1.96*pred_lt_1$se.fit,
           LL = pred_lt_1$fit - 1.96*pred_lt_1$se.fit,
           disc = ifelse(triceps$age<10, 1, 2)) %>%
  ggplot() + geom_point(data = triceps, aes(x = age, y = triceps), cex = .7, alpha = 0.2) +
  geom_line(aes(x = x, y = mean, group = disc), cex = 1.2) +
  geom_ribbon(aes(x = x, ymin = LL, ymax = UL), fill = "blue", alpha = 0.2) +
  theme_bw() -> p1
```

*Follow similar procedure for the 2 thresholds:*

```{r}
mod_lt_2 <- lm(triceps~age:(age<10)+age:(age<40), data = triceps)

predict(mod_lt_2, newdata = data.frame(age = triceps$age), se.fit = TRUE) -> pred_lt_2

data.frame(x = triceps$age,
           mean = pred_lt_2$fit,
           UL = pred_lt_2$fit + 1.96*pred_lt_2$se.fit,
           LL = pred_lt_2$fit - 1.96*pred_lt_2$se.fit,
           disc = case_when(triceps$age<10~1,
                            triceps$age>=10 & triceps$age<40~2,
                            triceps$age>=40~3)) %>%
  ggplot() + geom_point(data = triceps, aes(x = age, y = triceps), cex = .7, alpha = 0.2) +
  geom_line(aes(x = x, y = mean, group = disc), cex = 1.2) +
  geom_ribbon(aes(x = x, ymin = LL, ymax = UL), fill = "blue", alpha = 0.2) +
  theme_bw() -> p2
```


*plot together:*

```{r fig.height=4, fig.width=9}
p1|p2
```

*and the summary:*

```{r}
mod_lt_2 %>% summary()
mod_lt_2 %>% confint()
```


*For the model with the 3 thresholds: For every year increase in the age, for people aged less than 10 years old, there weak evidence of a 0.19 cm decrease in the triceps skinfold thickness, for people aged between 10 and 40 years old, for every year increase in the age, there is a 0.17 cm increase in the triceps skinfold thickness and for people 40 years old or older, for every year increase in the age, there is a 0.04 cm increase in the triceps skinfold thickness.*

*The discontinuity on the threshold is a limitation of this approach, because it does not allow us to investigate the natural progress of the process.*


Question 2.2
Now without using the \texttt{bs()} function or any kind of splines related packages in R, fit linear splines with one threshold at 10 years. Plot, calculate the 95% confidence intervals, interpret the results and compare with the results of the question 2.1.

*Reply:*
```{r fig.width=6, fig.height=4}
mod_lsplines <- lm(triceps~age + I((age-10)*(age<10)), data = triceps)

predict(mod_lsplines, newdata = data.frame(age = triceps$age), se.fit = TRUE) -> pred_lsplines

data.frame(x = triceps$age,
           mean = pred_lsplines$fit,
           UL = pred_lsplines$fit + 1.96*pred_lsplines$se.fit,
           LL = pred_lsplines$fit - 1.96*pred_lsplines$se.fit) %>%
  ggplot() + geom_point(data = triceps, aes(x = age, y = triceps), cex = .7, alpha = 0.2) +
  geom_line(aes(x = x, y = mean), cex = 1.2) +
  geom_ribbon(aes(x = x, ymin = LL, ymax = UL), fill = "blue", alpha = 0.2) +
  theme_bw()

mod_lsplines %>% summary()
mod_lsplines %>% confint()
```

*For people aged between 10 years or older for every year increase in the age there is a 0.26 cm increase in the triceps skinfold thickness. For every year increase in the age, for people aged less than 10 years old, there is a  -0.306+0.264 = -0.042 cm change in the in the triceps skinfold thickness*

Question 2.3
Using the \texttt{bs()} function, fit a 4 linear basis spline models in the data with one knot each time at 5, 10, 15 and plot all three fits. Which model is the most appropriate for our data?

*Reply:*
```{r fig.width=9, fig.height=4}
kn <- c(5, 10, 15)
list_res <- pred_res <- ci_graph <- list()
for(i in 1:3){
  list_res[[i]] <- lm(triceps~bs(age, degree=1, knots=kn[i]), data = triceps)
  pred_res[[i]] <- predict(list_res[[i]], newdata = data.frame(age = triceps$age), se.fit = TRUE)

  ci_graph[[i]] <- data.frame(
    est = pred_res[[i]]$fit,
    UL = pred_res[[i]]$fit + 1.96*pred_res[[i]]$se.fit,
    LL = pred_res[[i]]$fit - 1.96*pred_res[[i]]$se.fit,
    thr = kn[i]
  )
}

ci <- do.call(rbind, ci_graph)
ci$age <- rep(triceps$age, times = 3)
ggplot(data = ci) + geom_point(data = triceps, aes(x = age, y = triceps), cex = .7, alpha = 0.2) +
  geom_line(aes(x = age, y = est), cex = 1.2) +
  geom_ribbon(aes(x = age, ymin = LL, ymax = UL), fill = "blue", alpha = 0.2) +
  facet_grid(cols = vars(thr)) + theme_bw()
```

*The main problem with these models is that for different threshold they results a different fit. This threshold specific volatility makes them not very attractive to use when we have no information about the threshold. We can use criteria such as the waic to get an indication about better fits.*

```{r}
lapply(list_res, function(X){
  WAIC(X)$WAIC1 %>% return()
})
```

*with the smallest being for a knot at 5 years.*

Question 2.4
Take the \texttt{summary()} of the model from question 2.3 with a knot at 10 years and compare the results with the \texttt{lm()} fit of questions 2.2. What do you observe?

*Reply:*
```{r}
mod_lsplines %>% summary()
list_res[[2]] %>% summary()
```

*The coefficients are not the same.*

Question 2.5
In this question we will plot the basis function to understand the meaning of the coefficients of the \texttt{bs()} fit.
```{r message=FALSE, warning=FALSE, fig.width=9, fig.height=3}
b <- bs(triceps$age, degree = 1, knots = 10)
str(b)
b1 <- b[, 1]  ## basis 1
b2 <- b[, 2]  ## basis 2

data.frame(x=rep(triceps$age, times=2),
           b=c(b1, b2),
           basis=rep(c("basis 1", "basis 2"), each = triceps$age %>% length())) %>%
  ggplot() + geom_line(aes(x=x, y=b), cex=.8) + facet_grid(cols=vars(basis)) + theme_bw()
```

and now run the following:
```{r message=FALSE, warning=FALSE}
lm(triceps$triceps~b1+b2) %>% summary()
list_res[[2]] %>% summary()
```

What do the \texttt{ bs()} coefficients mean?

*Reply:The values of the coefficients in the bs version of the call are essentially scaling factors, that calibrate the basis function. What we are really interested in is the combination of those to get the overall fit.*

to get the estimates of the two separate lines:
```{r message=FALSE, warning=FALSE}
RegSplineAsPiecePoly(list_res[[2]], "bs(age, degree = 1, knots = kn[i])", shift = FALSE)
```

How can we interpret the results here? Compare the estimates with the ones retrieved in question 2.2.

*Reply:The interpretation now is the natural one without needing any reparametrisation: For every year increase in the age, for people aged less than 10 years old, there is a -0.0412 cm change in the in the triceps skinfold thickness. For every year increase in the age, for people older than 10 years old, there is a  0.265 cm increase in the in the triceps skinfold thickness.*

## Part 3: Triceps and subscapular skinfold thicknesses with splines

Question 3.1
Use the same dataset as before and similar with the question 2.5, use the function \texttt{bs()}  one knot at 10 and cubic splines

*Reply:*
```{r fig.width=9, fig.height=4}
b <- bs(triceps$age, degree = 3, knots = 10)
str(b)
b1 <- b[, 1]  ## basis 1
b2 <- b[, 2]  ## basis 2
b3 <- b[, 3]  ## basis 3
b4 <- b[, 4]  ## basis 4

data.frame(x=rep(triceps$age, times=4),
           b=c(b1, b2, b3, b4),
           basis=rep(c("basis 1", "basis 2", "basis 3", "basis 4"),
                     each = triceps$age %>% length())) %>%
  ggplot() + geom_line(aes(x=x, y=b), cex=.8) + facet_grid(cols=vars(basis)) + theme_bw()
```

Question 3.2
Use the \texttt{lm()} function ad \texttt{bs()} to fit cubic splines with one know at 10. What is the interpretation of the \texttt{summary()} output?

*Reply:*
```{r}
mod_bs_3_splines <- lm(triceps~bs(age, knots = 10, degree = 3), data = triceps)
mod_bs_3_splines %>% summary()
```

*The coefficients scale the evaluated B-spline basis functions according to the data in order to fit the curve.*

Question 3.3
Use the function \texttt{predict()} and plot the fit of the above model together with 95% confidence intervals.


*Reply:*
```{r fig.width=10, fig.height=4}
x2pred <- seq(from = 1, to = 50, length.out = 200)
predict(mod_bs_3_splines, newdata = data.frame(age=x2pred), se.fit = TRUE) -> pred_bs_3_splines

data.frame(
  x=x2pred,
  est=pred_bs_3_splines$fit,
  UL=pred_bs_3_splines$fit+1.96*pred_bs_3_splines$se.fit,
  LL=pred_bs_3_splines$fit-1.96*pred_bs_3_splines$se.fit
) %>%
  ggplot() + geom_point(data = triceps, aes(x = age, y = triceps), cex = .7, alpha = 0.2) +
  geom_line(aes(x = x, y = est), cex = 1.2) +
  geom_ribbon(aes(x = x, ymin = LL, ymax = UL), fill = "blue", alpha = 0.2) + theme_bw() -> p1
```


Question 3.4
Use natural splines instead and plot the fit of the above model together with 95% confidence intervals. Do you observed any differences? Tip: Include boundary knots in the \texttt{bs()} function.
```{r fig.width=9, fig.height=3}
mod_ns_3_splines <- lm(triceps~ns(age, knots = c(1, 10, 50)), data = triceps)
mod_ns_3_splines %>% summary()

predict(mod_ns_3_splines, newdata = data.frame(age=x2pred), se.fit = TRUE) -> pred_ns_3_splines

data.frame(
  x=x2pred,
  est=pred_ns_3_splines$fit,
  UL=pred_ns_3_splines$fit+1.96*pred_ns_3_splines$se.fit,
  LL=pred_ns_3_splines$fit-1.96*pred_ns_3_splines$se.fit
) %>%
  ggplot() + geom_point(data = triceps, aes(x = age, y = triceps), cex = .7, alpha = 0.2) +
  geom_line(aes(x = x, y = est), cex = 1.2) +
  geom_ribbon(aes(x = x, ymin = LL, ymax = UL), fill = "blue", alpha = 0.2) + theme_bw() -> p2


p1|p2
```

*The behavior of \texttt{ns()} in the boundaries is in general less wiggly, because it assumes linearity. Here it seems that the results are very similar.*

Question 3.5
Type \texttt{?ns} and check what the \texttt{df()} argument does. Now use only the argument \texttt{df()} in the \texttt{ns()} function and use the values 1, 10, 50, 100. What do you observe?
```{r eval=FALSE}
?ns
```

```{r}
lapply(c(1, 10, 50, 100), function(X){
  ggplot() + geom_point(data = triceps, aes(x = age, y = triceps), cex = .7, alpha = 0.2) +
    geom_line(aes(x = triceps$age, y = lm(triceps~ns(age, df=X),
                                          data = triceps)$fitted.values), cex = 1.2) +
    theme_bw() + ggtitle(paste("df=",X, sep = " ")) %>% return()
}) -> p_ns

(p_ns[[1]]|p_ns[[2]])/(p_ns[[3]]|p_ns[[4]])
```

*As we increase the degrees of freedom, them bias decreases, whereas as the degrees of freedom decrease the bias increases whereas the variance decreases.*

Question 3.6
Use the function \texttt{gam()} from the \texttt{mgcv} package and fit a penalized spline. What do you think about the fit?

*Reply:*

```{r}
mgcv::gam(triceps~s(age), family = "gaussian", data = triceps) -> mod_psplines
plot(mod_psplines)
```

or

```{r fig.width=6, fig.height=4}
predict (mod_psplines, data.frame(age = triceps$age), se.fit = TRUE) -> pred_splines
UL <- pred_splines$fit + 1.96*pred_splines$se.fit
LL <- pred_splines$fit - 1.96*pred_splines$se.fit

ggplot( ) +
  geom_point(data = triceps, aes(x=age, y=triceps), cex = 1) +
  geom_line(data = triceps, aes(x=age, y=pred_splines$fit) , linewidth = .5) +
  geom_ribbon(aes(ymin=LL, ymax=UL, x=triceps$age), fill="red", alpha=0.2) + theme_bw()
```

*It looks that the penalised spline provides a very good fit to the data, capturing all the main features/trends while at the same time not being very variable.*

## Part 4: Analysing the HANES dataset

In the fourth part of this practical you will be analyzing the HANES lb dataset. This dataset included information about the age, sex (men=0, women=1) race, location (locode), height, BMI (body mass index), Booze (categorical alcohol consumption), serum calcium (Ser.calc), setum cholesterol (Ser.chol), current smoking, smoking history, number of cigarettes per dat, lifetime pack year and follow up variables including age at death (d.age), year of death (d.year), death from any cause (d.total), death from cancer (d.cancer) and deaths from heart disease (d.heart)

As a first step we load the data and assign \texttt{hanes} to a data.frame.

```{r eval=TRUE}
hanes <- read.csv("../../data/hanes.csv")
```

Question 4.1
Fit a logistic regression model for mortality (d.total) with main effects for age, sex, race, booze, smokever, and bp1sys. Interpret the output of the model (focus on the coefficient of the systolic blood pressure).

*Reply:*
```{r}
glm(d.total~age+factor(sex)+factor(race)+factor(booze)+factor(smokever)+bp1sys,
    family = "binomial", data = hanes) -> mod1
mod1 %>% summary()
coefficients(mod1)["bp1sys"] %>% exp()
(mod1 %>% confint() %>% exp())*100
```

*After accounting for age, sex, ethnicity alcohol consumption and smoking we have observed a 0.8% (0.4-1.24%) increase in the odds for dying for every unit increase in the systolic blood pressure.*

Question 4.2
Now let's examine the interaction between systolic blood pressure and sex? Is it significant? Interpret the results.

*Reply:*
```{r}
glm(d.total~age+factor(sex)*bp1sys + factor(race)+factor(booze)+factor(smokever),
    family = "binomial", data = hanes) -> mod2
mod2 %>% summary()
```

*There is weak evidence of an interaction, implying that after accounting for age, sex, ethnicity alcohol consumption and smoking sex does not modify systolic blood pressure associated mortality risk.*

```{r}
anova(mod1, mod2)
```

*The likelihood ratio test also argues that there is weak evidence supporting an interaction term.*

Question 4.3
Repeat that analysis of question 3.1 using the \texttt{gam()} function and a smooth function for age. Is age nonlinear?
*Reply:*

```{r}
mgcv::gam(d.total~s(age)+factor(sex) + bp1sys + factor(race)+factor(booze)+factor(smokever),
          family = "binomial", data = hanes) -> mod3
mod3 %>% summary()

plot(mod3)
```

or

```{r fig.width=6, fig.height=4}
predict(mod3, hanes %>%
          dplyr::select(age, sex, bp1sys, race, booze, smokever) %>%
          mutate(sex=0, bp1sys=mean(bp1sys, na.rm = TRUE), race=1, booze=0, smokever=0),
        se.fit = TRUE) -> pred_mod3
UL <- pred_mod3$fit + 1.96*pred_mod3$se.fit
LL <- pred_mod3$fit - 1.96*pred_mod3$se.fit

ggplot() +
  geom_line(data=hanes, aes(x=age, y=pred_mod3$fit)) +
  geom_ribbon(data=hanes, aes(x=age, ymin=LL, ymax=UL) , fill=" blue", alpha =0.1) +
  theme_bw() + ylab("log odds")
```

*The effect looks reasonably linear.*

Question 4.4
Add a smooth function for \texttt{ser.chol}. Is it linear?

*Reply:*

```{r}
mgcv::gam(d.total~age+factor(sex) + bp1sys + factor(race)+factor(booze)+factor(smokever)+s(ser.chol),
          family = "binomial", data = hanes) -> mod4
mod4 %>% summary()
plot(mod4)
```

*It looks that the serum cholesterol needs a more flexible fit, but overall the effect seems to be around 0.*

Question 4.5
Now model d.heart instead of d.total. Is the dependence on \texttt{ser.chol} linear?

*Reply:*

```{r}
mgcv::gam(d.heart ~age+factor(sex) + bp1sys + factor(race)+factor(booze)+factor(smokever)+s(ser.chol),
          family = "binomial", data = hanes) -> mod5
mod5 %>% summary()
plot(mod5)
```

*The effect here seems to be linear. For parsirmonity reasons we will use a linear term instead.*

```{r}
mgcv::gam(d.heart ~age+factor(sex) + bp1sys + factor(race)+factor(booze)+factor(smokever)+ser.chol,
          family = "binomial", data = hanes) %>%
  summary()
```


## Part 5: Distribued lag linear models.

In the fifth part of this tutorial, we will use the \texttt{chicago} dataset of the \texttt{dlnm} package used throughout the lecture and examine the effect of heatwaves.

Question 5.1
What do you think a heatwave is? How should we define it?

*There is no universal definition of the heatwave. The three major components for defining a heatwave is a temperature metric (min, mean, max), the intensity (a threshold once reached implying extreme heat) and the duration (day for which the temperature with excess the above-mentioned threshold).*

Question 5.2
Let a heatwave be the period that we have observed temperatures higher that the 95th percentile of the overall temperature
in chicago for at least 2 consecutive days. Create a new variable called heatwave that takes values 0 if it falls within the
definition and 1 otherwise.

```{r message=FALSE, warning=FALSE}
library(dlnm)
chicagoNMMAPS %>% head()
```

You can use the following code to define the heatwave variable:

```{r message=FALSE, warning=FALSE}
threshold <- quantile(chicagoNMMAPS$temp, probs = 0.95)
chicagoNMMAPS %>%
  dplyr::mutate(
    heatwave = dplyr::case_when(temp >= threshold &
                                dplyr::lag(temp, 1) >= threshold &
                                dplyr::lag(temp, 2) >= threshold ~ 1,
                                TRUE ~ 0)
  ) -> chicagoNMMAPS

chicagoNMMAPS %>%
  dplyr::mutate(
    heatwave = dplyr::case_when(heatwave == 1 |
                                  dplyr::lead(heatwave, 1) == 1 |
                                  dplyr::lead(heatwave, 2) == 1 ~ 1,
                                TRUE ~ 0)
    ) -> chicagoNMMAPS
```

Quantify the effect of heatwave on resp deaths. Interpret the result. Hint: Include heatwave as linear, time as spline, month as spline and adjust for day of week and PM10

*Reply:*
```{r}
mgcv::gam(resp ~ s(time) + s(month) + dow + pm10 + factor(heatwave),
              data = chicagoNMMAPS , family = "poisson" ) -> mod_heatwave1

mod_heatwave1 %>% summary()

(exp(0.11 - 1.96*0.03) - 1)*100; (exp(0.11 + 1.96*0.03) - 1)*100
```

*There is a 5-18% increase in the respiratory mortality risk during heatwave days, after accounting for day of week, temporal trends and PM10.*

Question 5.3
Examine a potential lag effect of the heatwaves on respiratory mortality. Consider lags 1:10 and plot the effect. What do you observe?

*Reply:*

```{r}
k <- 1:10
res_store <- list()

for(i in 1:length(k)){

  if(i==1){
    chicagoNMMAPS %>%
      dplyr::mutate(
        heatwave_lagged = dplyr::case_when(heatwave == 1 |
                                             dplyr::lead(heatwave, 1) == 1 ~ 1,
                                           TRUE ~ 0)
      ) -> chicagoNMMAPS
  }else{
    chicagoNMMAPS %>%
      dplyr::mutate(
        heatwave_lagged = dplyr::case_when(heatwave_lagged == 1 |
                                             dplyr::lead(heatwave_lagged, 1) == 1 ~ 1,
                                           TRUE ~ 0)
      ) -> chicagoNMMAPS
  }


  mgcv::gam(resp ~ s(time) + s(month) + dow + pm10 + factor(heatwave_lagged),
            data = chicagoNMMAPS , family = "poisson" )  -> tmp

  res_store[[i]] <- list(est = (exp(coef(tmp)["factor(heatwave_lagged)1"])-1)*100,
                         LL = (exp(coef(tmp)["factor(heatwave_lagged)1"] -
                                     1.96*summary(tmp)$se["factor(heatwave_lagged)1"])-1)*100,
                         UL = (exp(coef(tmp)["factor(heatwave_lagged)1"] +
                                     1.96*summary(tmp)$se["factor(heatwave_lagged)1"])-1)*100)
}

lapply(res_store, unlist) %>% do.call(rbind, .) %>% as_tibble() %>%
  mutate(type =
           factor(paste0("lag ", k),
                  levels = paste0("lag ",k))) -> plotres

colnames(plotres)[1:3] <- c("est", "LL", "UL")

ggplot(data = plotres) +
  geom_point(aes(x=type, y=est)) +
  geom_errorbar(aes(x=type, ymin=LL, ymax=UL, width = 0.1)) +
  geom_hline(yintercept = 0, col = "red", linetype = "dotted") + theme_bw() +
  ylab("% increase in respiratory mortality") + xlab("Lags") -> p1


ggplot(data = plotres) +
  geom_point(aes(x=type, y=est)) +
  geom_line(aes(x=type, y=est, group=1)) +
  geom_ribbon(aes(x=type, ymin=LL, ymax=UL, group = 1), fill = "blue", alpha = 0.2) +
  geom_hline(yintercept = 0, col = "red", linetype = "dotted") + theme_bw() +
  ylab("% increase in respiratory mortality") + xlab("Lags") -> p2

p1/p2
```

*The effect of heatwave on the lag dimension looks linear. There evidence of an effect at lag 10 is weak.*

Question 5.4
Fit a distributed non-linear model. Consider a linear threshold model with the threshold being the 95th percentile of the temperature and lags 1:10.

```{r message=FALSE, warning=FALSE}
cb.heatwave <- crossbasis(chicagoNMMAPS$temp, lag=10,
                          argvar=list(fun="thr", thr = threshold),
                          arglag=list(fun="lin"))
```

Use the function \texttt{gam()} accounting for time as spline, month as spline and adjust for day of week and PM10. Take the \texttt{summary()} of the model and interpret the slopes corresponding to the cross basis function. What is the difference of this estimate with the heatwave estimate?
*Reply:*

```{r}
mgcv::gam(resp ~ s(time) + s(month) + dow + pm10 + cb.heatwave,
          data = chicagoNMMAPS, family = "poisson")  -> dlnm_heatwave_model

dlnm_heatwave_model %>% summary()
(exp(0.015)-1)*100
```


*Overall and across all lags considered there is a 1.5% increase in the respiratory mortality risk for every 1oC increase in temperatures higher than the 95th percentile of the temperature.*

*There is a declining trend of the effect of temperatures higher than the 95th percentile of the temperature across the lags with the log effect to reduce by -0.002 for every subsequent lag (0.2% decrease in the % respiratory mortality risk scale).*

*The heatwave estimate differs from this, as the heatwave shows a risk during heatwave days (a period of extreme heat), wheres the current estimate gives increase in risk for every 1$^o$C increase in the temperature.*

Question 5.5
Use the function \texttt{crosspred()}. Specify \texttt{at=-20:30} and \texttt{bylag=1}. Recall that \texttt{at} gives the temperature values and \texttt{bylag} the lags to predict. Retrieve the cumulative relative risk at 30$^o$C over the lags. What is the interpretation

*Reply:*

```{r}
pred.hw <- crosspred(cb.heatwave, dlnm_heatwave_model, at=-20:30, bylag=1)

pred.hw$allRRfit["30"]
pred.hw$allRRlow["30"]
pred.hw$allRRhigh["30"]
```

*Relative to temperatures lower than the 95th percentile of the temperature, there is a 20% (2%, 43%) increased respiratory mortality risk across the 0-10 lags when the temperature is 30$o$C.*

Question 5.6
Plot a 3D plot to show the lag, temperature and relative risk dimension. Also, plot 2 slices of the 3D plot, one for lag=1 and the second for temperatures=30. If interested, you can use the \texttt{plotly} package to produce an interactive 3D plot.

```{r}
plot(pred.hw, theta=150, phi=10, lphi=100, xlab = "Temperature (\u00B0C)", zlab="RR")
```


```{r fig.width=5, fig.height=4}
plot(pred.hw, "slice", lag=1, las = 1,
     main = "Relative respiratory mortality risk: lag 1",
     xlab = "Temperature (\u00B0C)", ylab = "", cex.main=1.0)
plot(pred.hw, "slice", var=30, las = 1,
     main = "Relative respiratory mortality risk: Temperature 30\u00B0C",
     xlab = "Lag", ylab = "", cex.main=1.0)
```

*and with the \texttt{plotly} package:*

```{r}
p <- plot_ly()
p <- add_surface(p,
               x = 0:10, # the lags
               y = -20:30, # the temperature as is on the pred.hw$matRRfit defined by at
               z = pred.hw$matRRfit)

layout(p, scene = list(xaxis = list(title = "Lag"),
                      yaxis = list(title = "Temperature (\u00B0C)", range = c(0,30)),
                      zaxis = list(title = "Relative risk"))
)
```


<!--
require(knitr)
require(markdown)
require(rmarkdown)
render("practical1-lm-glm-re.Rmd")
-->

